---
output:
  md_document:
    variant: markdown_github
bibliography: README.bib
csl: inst/bib_style.csl
---

# pre: an R package for deriving prediction rule ensembles

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/README-figures/README-",
  dpi = 124
)
```

pre is an R package for deriving prediction rule ensembles for binary and continuous outcome variables. Input variables may be numeric, ordinal and nominal. The package implements the algorithm for deriving prediction rule ensembles as described in [@Friedman08], with several adjustments: 

1) The package is completely R based, allowing users better accessible results and more control over the parameters used for generating the prediction rule ensemble.
2) The unbiased tree induction algorithm of [@Hothorn06] is used for deriving prediction rules, instead of the classification and regression tree (CART) algorithm, which suffers from biased variable selection.
3) The package allows for plotting the final rule ensemble as a collection of simple decision trees.
4) The initial ensemble of prediction rules can be generated as a bagged, boosted and/or random forest ensemble.
5) Hinge functions of predictor variables may be included as baselearners in the ensemble, as in the multivariate adaptive regression splines technique of [@Friedman91].

The pre package is developed to provide useRs a completely R based implementation of the algorithm described by [@Friedman08]. However, note that pre is under development, and much work still needs to be done. See [@Fokkema15] for an application of the methods.

# Examples

To get a first impression of how pre works, we will fit a prediction rule ensemble to predict Ozone levels using the airquality dataset. WE can fit a prediction rule ensemble using the pre() function:

```{r, results = FALSE}
library(pre)
set.seed(42)
airq.ens <- pre(Ozone ~ ., data = airquality[complete.cases(airquality), ])
```

We can print the resulting ensemble using the print function(): 

```{r}
print(airq.ens)
```

We can plot the rules and linear terms in the ensemble using the plot() function:

```{r, fig.height=6, fig.width=6}
plot(airq.ens, penalty.par.val = "lambda.1se", max.terms.plot = 9, cex = .6)
```

We can obtain the estimated coefficients for each of the baselearners using the coef() function:

```{r}
coefs <- coef(airq.ens)
coefs[1:6,]
```

We can assess the importance of the input variables as well as the baselearners in the ensemble using the importance() function:

```{r, fig.height=4, fig.width=4}
importance(airq.ens, round = 4)
```

We can generate predictions for new observations using the predict() function:

```{r}
airq.preds <- predict(airq.ens, newdata = airquality[1:4,])
```

We can obtain a partial dependence plot to assess the effect of a single predictor variable on the outcome using the singleplot() function:

```{r, fig.width=5, fig.height=3}
singleplot(airq.ens, varname = "Temp")
```

We can obtain a partial dependence plot to assess the effect of a pair of predictor variables on the outcome using the pairplot() function:

```{r,  fig.width=5, fig.height=5}
pairplot(airq.ens, varnames = c("Temp", "Wind"))
```

We can assess the expected prediction error of the ensemble through cross validation (10-fold, by default) using the cvpre() function:

```{r}
set.seed(43)
airq.cv <- cvpre(airq.ens)
airq.cv$accuracy
```

More complex prediction ensembles can be obtained using the gpe() function. The abbreviation gpe stands for generalized prediction ensembles, which may include hinge functions of the predictor variables as described in [@Friedman91], in addition to rules and/or linear terms. Addition of such hinge functions may improve predictive accuracy (but may also reduce interpretability).  

# References