---
output:
  md_document:
    variant: markdown_github
bibliography: README.bib
csl: inst/bib_style.csl
---

# pre: An R package for deriving prediction rule ensembles

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/README-figures/README-",
  dpi = 124
)
```

pre is an R package for deriving prediction rule ensembles for binary and continuous outcome variables. Input variables may be numeric, ordinal and nominal. The package implements the algorithm for deriving prediction rule ensembles as described in [@Friedman08], with some improvements and adjustments. The most important improvements and adjustments are: 

1) The package is completely R based, allowing users better accessible results and more control over the parameters used for generating the prediction rule ensemble
2) The unbiased tree induction algorithm of [@Hothorn06] is used for deriving prediction rules, instead of the classification and regression tree (CART) algorithm, which suffers from biased variable selection.
3) The package allows for plotting the final rule ensemble as a collection of simple decision trees.
4) The initial ensemble of prediction rules can be generated as a bagged, boosted and/or random forest ensemble.
5) Hinge functions of predictor variables may be included as baselearners in the ensemble, as in the multivariate adaptive regression splines technique of [@Friedman91].

The pre package is developed to provide useRs a completely R based implementation of the algorithm described by [@Friedman08]. However, note that pre is under development, and much work still needs to be done. See [@Fokkema15] for an application of the methods.

# Examples

To get a first impression of how pre works, we will fit a prediction rule ensemble to predict Ozone levels using the airquality dataset:

```{r, results = FALSE}
library(pre)
complete <- complete.cases(airquality)
set.seed(42)
airq.ens <- pre(identity(Ozone + Temp) ~ ., data = airquality[complete, ], standardize = TRUE)
```

We can print the resulting ensemble: 

```{r}
print(airq.ens)
```

We can plot a subsample of the rules (and or/linear terms) in the ensemble:

```{r, fig.height=8, fig.width=8}
plot(airq.ens, max.terms.plot = 9, cex = .75, penalty = "lambda.1se")
```

We can obtain the estimated coefficients of the ensemble:

```{r}
head(coef(airq.ens))
```

We can assess the importance of the predictor variables, and each of the rules and/or linear terms in the ensemble:

```{r}
importance(airq.ens)
```

We can generate predictions for new observations:

```{r}
airq.preds <- predict(airq.ens, newdata = airquality[1:4,])
```

We can assess the effect of a given predictor variable on the outcome through a partial dependence plot:

```{r}
singleplot(airq.ens, varname = "Temp")
```

We can assess the expected prediction error of the ensemble, by default calculated using 10-fold cross validation:

```{r}
set.seed(43)
airq.cv <- cvpre(airq.ens)
airq.cv$accuracy
```

More complex prediction ensembles can be derived with the gpe() function. The abbreviation gpe stands for generalized prediction ensembles, which in addition to rules and linear terms may also include hinge functions of the predictor variables [@Friedman91]. Addition of hinge functions may improve predictive accuracy (but may also reduce interpretability).  

# References