---
title: "Dealing with missing data in fitting prediction rule ensembles"
author: "Marjolein Fokkema"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dealing with missing data in fitting prediction rule ensembles}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## Introduction

To deal with missing data, multiple imputation is the golden standard (Schafer & Graham, 2002). With GLMs, the models fitted on each imputed dataset can then be pooled. For non-parameteric methods like prediction rule ensembles, such pooling is more difficult. Little research has been performed on how to best deal with missing data in fitting prediction rule ensembles, but there are currently three options:

* Listwise deletion. Although the default in `pre()`, this is certainly the least favorable option.

* Single imputation: Perform only a single imputation and fit a prediction rule ensemble on this single dataset. This is likely better than listwise deletion, but likely inferior to multiple imputation, but easy to implement.

* Multiple imputation approach by aggregating ensembles: Create multiple imputed datasets; fit a separate prediction rule ensemble to each of the imputed datasets; aggregate the ensembles into a single final ensemble. In terms of predictive accuracy, this approach will work well. It will, however, yield more complex ensembles than the former two approaches, and the next approach.

* Combining mean imputation with the Missing-In-Attributes approach. According to Josse et al. (2019), mean imputation and the Missing-In-Attributes approaches perform well from a prediction perspective. Furthermore, they are computationally inexpensive. 

Below, we provide examples for the first three approaches described above. In future versions of package **`pre`**, the mean imputation combined with MIA approach will be implemented.



## Example: Predicting wind speed

For the examples, we will be predicting Wind speels using the `airquality` dataset (we focus on predicting the `wind` variable, because it does not have missing values,  while variables `Ozone` and `Solar.R` do):

```{r, results='hide', message=FALSE, warning=FALSE, fig.width=5}
head(airquality)
nrow(airquality)
library("mice")
md.pattern(airquality, rotate.names = TRUE)
```

## Listwise Deletion

This option is the default of function `pre()`:

```{r}
library("pre")
set.seed(43)
airq.ens <- pre(Wind ~., data = airquality)
airq.ens
```
With listwise deletion, only `r sum(complete.cases(airquality))` out of `r nrow(airquality)` observations are retained. We obtain a rather sparse ensemble.


## Single Imputation

Here we apply single imputation by replacing missing values with the mean: 

```{r}
imp0 <- airquality
imp0$Solar.R[is.na(imp0$Solar.R)] <- mean(imp0$Solar.R, na.rm=TRUE)
imp0$Ozone[is.na(imp0$Ozone)] <- mean(imp0$Ozone, na.rm=TRUE)
set.seed(43)
airq.ens.imp0 <- pre(Wind ~., data = imp0)
airq.ens.imp0
```

We obtain a larger number of rules, and slightly lower cross-validated mean squared error. However, this model cannot really be compared with the listwise deletion model, because they are computed over different sets of observations.


## Multiple Imputation

We perform multiple imputation by chained equations, using the predictive mean matching method. We generate 5 imputed datasets:

```{r, results='hide', message=FALSE, warning=FALSE, fig.width=7}
set.seed(42)
imp <- mice(airquality, m = 5)
```

We create a `list` with imputed datasets:

```{r}
imp1 <- complete(imp, action = "all", include = FALSE)
```

We load the **`pre`** library:

```{r}
library("pre")
```

We create a custom function that fits PREs to several datasets contained in a list:

```{r}
pre.agg <- function(datasets, ...) {
  result <- list()
  for (i in 1:length(datasets)) {
    result[[i]] <- pre(datasets[[i]], ...)
  }
  result
}
```

We apply the new function:

```{r}
set.seed(43)
airq.agg <- pre.agg(imp1, formula = Wind ~ .)
```

Note that we can used the ellipsis (`...`) to pass arguments to `pre` (see `?pre` for an overview of arguments that can be specified).

We now define `print`, `summary`, `predict` and `coef` methods to extract results from the fitted ensemble. Again, we can use the ellipsis (`...`) to pass arguments to the `print`, `summary`, `predict` and `coef` methods of function `pre` (see e.g., `?pre:::print.pre` for more info):

```{r, results ='hide'}
print.agg <- function(object, ...) {
  result <- list()
  sink("NULL")
  for (i in 1:length(object)) {
    result[[i]] <- print(object[[i]], ...)
  }
  sink()
  print(result)
}
print.agg(airq.agg) ## results suppressed for space considerations

summary.agg <- function(object, ...) {
  for (i in 1:length(object)) summary(object[[i]], ...)
}
summary.agg(airq.agg) ## results suppressed for space considerations
```

For averaging over predictions, there is only one option for continuous outcomes. For non-continuous outcomes, we can average over the linear predictor, or over the predicted values on the scale of the response. I am not sure which would be more appropriate; the resulting predicted values will not be identical, but highly correlated, though.

```{r}
predict.agg <- function(object, newdata, ...) {
  rowMeans(sapply(object, predict, newdata = newdata, ...))
}
agg_preds <- predict.agg(airq.agg, newdata = airquality[1:4, ])
agg_preds
```

Finally, the `coef` method should return the averaged / aggregated final PRE. That is, it returns: 

1) One averaged intercept; 

2) All rules and linear terms, with their coefficients scaled by the number of datasets; 

3) In presence of identical rules and linear terms, it aggregates those rules and their coefficients into one rule / term, and adds together the scaled coefficients. 

Note that linear terms that do not have the same winsorizing points will not be aggregated. Note that the labels of rules and variables may overlap between different datasets (e.g., the label `rule 12` may appear multiple times in the aggregated ensemble, but each `rule 12` will have different conditions).

```{r}
coef.agg <- function(object, ...) {
  coefs <- coef(object[[1]], ...)
  coefs <- coefs[coefs$coefficient != 0, ]
  for (i in 2:length(object)) {
    coefs_tmp <- coef(object[[i]], ...)
    coefs <- rbind(coefs, coefs_tmp[coefs_tmp$coefficient != 0, ])
  }
  ## Divide coefficients by the number of datasets:
  coefs$coefficient <- coefs$coefficient / length(object)
  ## Identify identical rules:
  duplicates <- which(duplicated(coefs$description))
  for (i in duplicates) {
    first_match <- which(coefs$description == coefs$description[i])[1]
    ## Add the coefficients:
    coefs$coefficient[first_match] <- 
      coefs$coefficient[first_match] + coefs$coefficient[i]
  }
  ## Remove duplicates:
  coefs <- coefs[-duplicates, ]
  ## Check if there are- duplicate linear terms left and repeat:
  duplicates <- which(duplicated(coefs$rule))
  for (i in duplicates) {
    first_match <- which(coefs$rule == coefs$rule[i])[1]
    coefs$coefficient[first_match] <- 
      coefs$coefficient[first_match] + coefs$coefficient[i]
  }
  coefs <- coefs[-duplicates, ]
  ## Return results:
  coefs
}
coef.agg(airq.agg)
```

We have obtained a final ensemble of `r nrow(coef.agg(airq.agg))-1` terms.




## Comparing accuracy and sparsity

We compare performance using 10-fold cross validation. We evaluate predictive accuracy and the number of selected rules. We only evaluate accuracy for observations that have no missing values.

```{r, eval=FALSE, results='hide', message = FALSE, warning = FALSE}
k <- 10
set.seed(43)
fold_ids <- sample(1:k, size = nrow(airquality), replace = TRUE)

observed <- c()
for (i in 1:k) {
  ## Separate training and test data
  test <- airquality[fold_ids == i, ]
  test <- test[!is.na(test$Ozone), ]
  test <- test[!is.na(test$Solar.R), ]
  observed <- c(observed, test$Wind)
}  

preds <- data.frame(observed)
preds$LWD <- preds$SI <- preds$MI <- preds$observed
nterms <- matrix(nrow = k, ncol = 3)
colnames(nterms) <- c("LWD", "SI", "MI")
row <- 1

for (i in 1:k) {

  if (i > 1) row <- row + nrow(test)
  
  ## Separate training and test data
  train <- airquality[fold_ids != i, ]
  test <- airquality[fold_ids == i, ]
  test <- test[!is.na(test$Ozone), ]
  test <- test[!is.na(test$Solar.R), ]

  ## Fit and evaluate listwise deletion
  premod <- pre(Wind ~ ., data = train)
  preds$LWD[row:(row+nrow(test)-1)] <- predict(premod, newdata = test)
  tmp <- print(premod)
  nterms[i, "LWD"] <- nrow(tmp) - 1
  
  ## Fit and evaluate single imputation
  imp0 <- train
  imp0$Solar.R[is.na(imp0$Solar.R)] <- mean(imp0$Solar.R, na.rm=TRUE)
  imp0$Ozone[is.na(imp0$Ozone)] <- mean(imp0$Ozone, na.rm=TRUE)
  premod.imp0 <- pre(Wind ~., data = imp0)
  imp1 <- test
  imp1$Solar.R[is.na(imp1$Solar.R)] <- mean(imp0$Solar.R, na.rm=TRUE)
  imp1$Ozone[is.na(imp1$Ozone)] <- mean(imp0$Ozone, na.rm=TRUE)
  preds$SI[row:(row+nrow(test)-1)] <- predict(premod.imp0, newdata = imp1)  
  tmp <- print(premod.imp0)
  nterms[i, "SI"] <- nrow(tmp) - 1
  
  ## Perform multiple imputation
  imp <- mice(train, m = 5)
  imp1 <- complete(imp, action = "all", include = FALSE)
  airq.agg <- pre.agg(imp1, formula = Wind ~ .)
  preds$MI[row:(row+nrow(test)-1)] <- predict.agg(airq.agg, newdata = test)
  nterms[i, "MI"] <- nrow(coef.agg(airq.agg)) - 1

}
```


```{r, echo=FALSE, eval=FALSE}
save(preds, nterms, file = "Missing_data_results.Rda")
```

```{r, echo=FALSE}
load("Missing_data_results.Rda")
```

```{r, fig.width=5, fig.height=5}
sapply(preds, function(x) mean((preds$observed - x)^2)) ## MSE
sapply(preds, function(x) sd((preds$observed - x)^2)/sqrt(nrow(preds))) ## SE of MSE
var(preds$observed) ## benchmark: Predict mean for all
```

Interestingly, we see that all three methods yield similar predictions and accuracy, and explain about 20% of variance in the response. Multiple imputation performed best, followed by listwise deletion, followed by single imputation. Taking into account the standard errors, however, these differences are not significant. Also, this simple evaluation on only a single dataset should not be taken too seriously. The better performance of multiple imputation does come at the cost of increased complexity: 

```{r, fig.width=4, fig.height=4}
boxplot(nterms, main = "Number of selected terms \nper missing-data method",
        cex.main = .8)
```

In line with findings of Josse et al. (2019), we expect MIA to work better for rules than mean imputation. In future versions of package **`pre`**, we plan to implement MIA (for the rules) and combine it with mean imputation (for the linear terms).

## Session info

In case you obtained different results, these results were obtained using the following:

```{r}
sessionInfo()
```

## References

Josse, J., Prost, N., Scornet, E., & Varoquaux, G. (2019). On the consistency of supervised learning with missing values. *arXiv preprint arXiv:1902.06931*. https://arxiv.org/abs/1902.06931

Miles, A. (2016). Obtaining predictions from models fit to multiply imputed data. *Sociological Methods & Research, 45*(1), 175-185. https://doi.org/10.1177/0049124115610345

Schafer, J. L., & Graham, J. W. (2002). Missing data: our view of the state of the art. *Psychological Methods, 7*(2), 147. https://doi.org/10.1037/1082-989X.7.2.147


```{r, echo = FALSE}
# Austin et al. (2019) refer to Wood et al. (2008) for stacking with multiply imputed data:
# 
# "Stacked Imputed Datasets With Weighted Regressions (W1, W2, and W3)
# 
# This approach entails stacking the M imputed datasets into 1 large dataset and then conducting variable selection in this single stacked dataset. To account for the multiple observations for each subject, weights are incorporated into the regression model when conducting variable selection. Wood et al proposed 3 different sets of weights that could be used: W1: w=1/M, in which each subject is weighted using the reciprocal of the number of imputed datasets; W2: w=(1−f)/M, where f denotes the proportion of missing data across all variables; W3: wj=(1−fj)/M, where fj denotes the proportion of missing data for variable Xj. Using the third approach, a different set of weights is used when assessing the statistical significance of a given candidate predictor variable."
# 
# Austin, P. C., Lee, D. S., Ko, D. T., & White, I. R. (2019). Effect of variable selection strategy on the performance of prognostic models when using multiple imputation. Circulation: Cardiovascular Quality and Outcomes, 12(11), e005927.
# 
# Wood, A. M., White, I. R., & Royston, P. (2008). How should variable selection be performed with multiply imputed data?. Statistics in medicine, 27(17), 3227-3246.

## Does glmnet allow for such an approach? Seems not, as weights are scaled automatically, see also https://stats.stackexchange.com/a/196615/173546
```